---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "Christian Andriolas"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

In the class activity, you tested a school board member's claim using data from Montgomery County, Pa. — and found a model that didn't hold up. Now you'll run the same analysis using data from Montgomery County, Md. Your job is to evaluate whether the Maryland data tells a different story, and what that means for how you'd report on the claim.

### Loading the Data

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```

### Task 1: Examine the Data First (2 points)

Before you plot anything, look at the data — you know why by now.

```{r}
summary(schools$`FARMS Pct`)
schools |> arrange(desc(`FARMS Pct`)) |> head(10)
```

Are there any data quality issues like the ones you found in the Pennsylvania data? How does the distribution of FARMS percentages compare between the two counties? Note anything that concerns you or gives you confidence in the data.

Not having so many schools with a FARMS percentage of 100% takes away the data quality issues. This allows the data to be spread out than the PA data which is a better comparison among the schools. 


### Task 2: Create a Scatterplot and Compare (5 points)

Replace the REPLACE_ME value with the correct variable name to create a scatterplot of FARMS percentage vs. math scores for Montgomery County, Md.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage, Montgomery County MD")
```

Describe what you see. Then compare this scatterplot specifically to the Pennsylvania one: what differences do you notice in the direction, strength, and spread of the relationship? The two counties share a name and are both suburban — what might explain why they look so different?

The scatterplot shows a negative correlations between FARMS percentage increasing and the students' scores on math exams. The scatterplots might look different because Montgomery County, MD is one of the wealthiest counties in the country so students are less likely to need FARMS, cuasing the plot to look different. 


### Task 3: Linear Regression Model (5 points)

Replace the REPLACE_ME value with the correct variable name to run a linear regression model predicting math scores from FARMS percentage.

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)
summary(model)
```

Report the three key statistics: the p-value, the Adjusted R-squared, and the coefficient for FARMS Pct. Then interpret them together — not separately. What does the combination of these three numbers tell you about the relationship between poverty and math scores in Montgomery County, Md.? Is this model worth using? How does it compare to the Pennsylvania model?

The p-vlaue is 0.0000000000000002. The adjusted r-squared value is 6734 and the coefficient value is -0.59511. These values prove their is a correlation between the variables and the data is statistically significant. The negative number means we can reject the null hypothesis. I think the model is worth using because it confirms a negative correlation and that the values don't effect each other. It is more significant than the PA model it provides a conclusive answer which the PA model never did. 


### Task 4: Identifying Schools with the Biggest Residuals (6 points)

Replace the REPLACE_ME values with the correct variable names to calculate predicted values, residuals, and absolute residuals for each school. Then arrange the schools in descending order of absolute residuals.

```{r}
schools <- schools |>
  mutate(
    predicted = predict(model),
    residual = `% Met or Exceeded Math Expectations` - predicted,
    abs_residual = abs(residual)
  ) |>
  arrange(desc(abs_residual))

head(schools |> select(1, `FARMS Pct`, `% Met or Exceeded Math Expectations`, predicted, residual), 10)
```

Identify the school with the largest positive residual (doing much better than the model predicts) and the school with the largest negative residual (doing much worse). For each school, find its zip code and look it up on [CensusReporter.org](https://censusreporter.org).

For each of the two schools, answer:
- What does the demographic and economic profile of the zip code look like?
- Does anything in that profile help explain why the school is over- or under-performing the model?
- What would your first reporting question be to the principal of that school?

School with the largest positive residual: Fox Chapel Elementary. The population is 26,406 with a lower average income and 9.8% of persons below the poverty line. 51% male and 49% female with an average age of 36.8 with 23% under 18 and a diverse population. 
School with the largest negative residual: Germantown Elementary. The population of the area code is 62,738 and it has a higher average income with 8.3% of persons below the poverty line. 51% male and 49% female with an average age of 37.6 with 23% under 18 and a diverse population. 
I don't see a glaring reason why either school is over or underperforming. The one factor that makes the most sense to me is the zip code for Germantown elementary having far more people which might leave less resources for teachers to help individual students and lead to a larger negative residual. I would first ask the Fox Chapel principal asking what they thin is the reason students are overperforming based on this metric. I would ask the Germantown principal how the school can create better resources to help individual students better. 




### Task 5: Evaluate the Model and Write the Pitch (7 points)

You've now run this analysis on two counties. Return to the school board member's claim: that elementary schools with fewer students on free and reduced-price meals do better on standardized math tests.

Write a short story pitch of 4-6 sentences that addresses all of the following: Does the Maryland data support or contradict the claim? How does it compare to what you found in Pennsylvania? What would you ask the school board member if you called them for comment? And what's the most interesting specific finding from your residuals analysis: which school would you want to visit first, and why?

The school board members claim that higher poverty levels correlate with lower test scores measured by FARMS percentage is strongly supported by the data in Montgomery County, MD. The PA data quality issues and lack of significance made it hard to support the claim, but the data in the Montgomery County, MD model is much more reliable. My question to the school board member would be what they think is the reason for such different data between the two counties. The most interesting finding is that Fox Chapel Elementary significantly outperformed the model's predictions. 
