---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2025.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

ANSWER HERE: What I notice about the data is there employees and detailee's that work in the white house, but there are many more employees. When scrolling through the data, it would be many people listed as an employee then one listed as a detailee and that trend continued throughout. Eight workers don't earn a salary and are all listed as employees. Each one of them has either a title like advisor, senior, czar, or the word "president" in their job description. It was also interesting that many employees had a salary of $195,200 but and only one had one higher than that and it was much higher at $225,700 as a detailee. From what I notice, I don't think any salary values will cause problems for analysis. 

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |> summarize(mean_salary = mean(SALARY),median_salary = median(SALARY))
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |> summarize(mean_salary = mean(SALARY),
    median_salary = median(SALARY),
    min_salary = min(SALARY),
    max_salary = max(SALARY))
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

ANSWER HERE: The mean and median salary being close in values shows that the data is pretty symmetrical. Outliers such as the few employees no salaries at the minumum value and the one detialee the maximum salary value of over $30,000 higher than anybody else didn't cause a lot of variance in the numbers. I would count the $0 minimum salary as an outlier and the $225,700 salary to be close but not considered an outlier becaise there are many other empluyees that have $30,000 differences in salaries like the difference between the second largst and max salary, but the $0 salary is even near any other value. 

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

ANSWER HERE: A story idea for me would be why is the max salary $30,000 higher than the next group of salaries that many employees have? I would try to understand why this detailee's job pays so much higher and if there is data to prove why they receive much more than anybody else. Their title is listed as an advisor, so I'd find the mean and median salary for all workers listed as an advisor without the max salary, then do the same thing with the max salary to see the difference it makes. 
